Why do we Eval?\\[2mm]\Large A large-scale study of Eval usage in R

### 2021-01-03 paper/main.tex
Fix number in abstract
Update 3.2

\begin{abstract}
  \noindent Most dynamic languages allow users to turn text into code
  using various functions, often named \eval, with language-dependent
  semantics. The widespread use of these reflective functions hinders
  static analysis and prevents compilers from performing
  optimizations. This paper aims to provide a better sense of why
  programmers use \eval. Understanding why \eval is used in practice
  is key to finding ways to mitigate its negative impact. We have
  reasons to believe that reflective feature usage is language and
  application domain specific; we focus on data science code written
  in R, and compare our results to previous work that analyzed web
  programming in JavaScript. Dynamic analysis of a corpus of
  \CorpusCodeRnd lines of libraries and end-user code confirms that
  \eval is indeed in widespread use; R's \eval is more dangerous in
  some ways, and safer in others, than what was previously reported
  for JavaScript.
  %% TODO ::: Add a sentance about summarizing why users resort to
  %% eval -- Jan
\end{abstract}

@@

Our corpus is assembled out of two set of programs, one consisting of libraries
obtained from \emph{CRAN} (Comprehensive R Archive Network\footnote{CRAN is the
  largest repository of R code with almost \CranAvailablePackagesRnd packages.
  It receives about 6 new package submissions a day~\cite{Ligges2017}}) and the
other from the \emph{Kaggle} data science competition website\footnote{\cf
  \url{https://kaggle.com}}. The intent here is to contrast code written by
experienced developers (CRAN) with code authored by typical end users of the
language (Kaggle). Another difference is that CRAN is a curated repository. Each
submitted package must abide to a number of well-formedness rules that are
automatically checked asserting certain quality. Most relevant for this work is
that all of the runnable code is tested and only a successfully running package
is admitted in the archive. Kaggle on the other hand is more like GitHub, anyone
can submit any code. All the code in our corpus is runnable, \ie we have scripts
that invoke the CRAN packages and the Kaggle programs have input data.

@@

The results presented in this paper are the result of an automated
analysis pipeline that acquires the code of packages, extract
metadata, executes programs, traces their behavior and summarizes the
observations. Figure~\ref{fig:pipeline} shows the main steps of the
pipeline along with approximate time to execute each step, the data
size, and the number of elements manipulated by the stage. Timings are
for runs on an Intel Xeon 6140, 2.30GHz with 72 cores and 256GB of
RAM.

\begin{figure}[!h]\hspace{-5mm}
  \includegraphics[width=1.05\linewidth]{pipeline.pdf}
  \caption{Analysis Pipeline}\label{fig:pipeline}
\end{figure}

The first step in the pipeline consist of downloading CRAN packages
along with their dependencies and acquiring Kaggle programs with the
help of a web crawler. The second step, is to compute code size and
coverage metrics for the CRAN packages (we use the \covr package for
coverage). The third step consist in extracting runnable programs from
packages: i.e. the tests, examples and vignettes. Each extracted
program is wrapped into a call to our dynamic analyzer --- the tool is
called \emph{evil} for \underline{ev}al \underline{i}nspection
\underline{l}ibrary. This step is needed to ensure that we record
\eval usage only for the target package. Without this, the data would
include eval calls from the unit testing frameworks as well as from
bootstrapping R virtual machine itself. To avoid any interference,
each program is run in its own R instance.
The fourth step in the pipeline is to perform dynamic analysis for
each run of a CRAN package or Kaggle program.

The dynamic analyzer builds upon the dynamic analysis framework,
\instrumentr that we have implemented to enable us to write dynamic
analysis logic in R. \instrumentr serves as an intermediary between
\rdyntrace and \evil, it intercepts the hooks exposed by \rdyntrace
and attaches R functions exported by \evil as callbacks. The \evil
callbacks execute on corresponding interpreter events.

The data extracted by \evil from each program is
concatenated, cleaned and summarized in the post-processing phase by
custom R scripts. Finally, the summarized data is analyzed in
RMarkdown notebooks to gather insights. Apart from the figures, the
data points included in the paper are also generated by RMarkdown
notebooks as latex macros.

The \emph{evil} framework is implemented as a R package in 2K lines of
R and 400 lines of C++ code. \instrumentr is an R package implemented
in 2.5K lines of R and 6K lines of C++ code. It internally uses a
modified R interpreter, \rdyntrace~\cite{oopsla19a}, that exposes
hooks from within the interpreter implementation for events of
interest.

All steps of this pipeline are parallelized using GNU
parallel~\cite{GNUparallel} and orchestrated by GNU make. To schedule
and parallelize extraction and analysis of programs, we use the \runr
package. Furthermore, \runr gracefully handles and reports failures
across large-scale program runs which greatly aids debugging of the
analysis pipeline.

\mypara{Limitations.} Dynamic analysis can only observe
calls that are triggered by the program's input. We believe that
focusing on R packages with high code coverage does mitigate this to
some extent. The results we report here were obtained with R's
bytecode compiler turned off, this should not affect the results as
the compiler does not optimize \eval.

We turn off the bytecode compiler for this study. The bytecode
compiler can also call \eval. We do not get source locations for
\UndefinedEvalsRnd \eval calls. In these cases \eval is either passed
as an argument to a higher-order functions or is defined in a function
returned by a higher-order function and the R parser does not retain
location information for \eval. However, this is a meager
\PercentUndefinedEval of all \eval calls and is unlikely to affect our
analysis. We ignore calls to the native \eval function exposed by R.
We also ignore the \c{rlang::tidy_eval} function which uses native
\eval internally because \c{rlang} is used to implement a DSL for data
analysis in R. It introduces a new first-class promise object called
\c{quosure} for which it implements special evaluation support in
\c{tidy_eval}.

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Usage Metrics}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\begin{abstract}
  \noindent Most dynamic languages allow users to turn text into code
  using various functions, often named \eval, with language-dependent
  semantics. The widespread use of these reflective functions hinders
  static analysis and prevents compilers from performing
  optimizations. This paper aims to provide a better sense of why
  programmers use \eval. Understanding why \eval is used in practice
  is key to finding ways to mitigate its negative impact. We have
  reasons to believe that reflective feature usage is language and
  application domain specific; we focus on data science code written
  in R, and compare our results to previous work that analyzed web
  programming in JavaScript. Dynamic analysis of a corpus of
  \CranAllPackages libraries and \KaggleCode end-user code confirms that
  \eval is indeed in widespread use; R's \eval is more dangerous in
  some ways, and safer in others, than what was previously reported
  for JavaScript.
  %% TODO ::: Add a sentance about summarizing why users resort to
  %% eval -- Jan
\end{abstract}

@@

Our corpus is assembled out of two set of programs, one consisting of libraries
obtained from \emph{CRAN} (Comprehensive R Archive Network\footnote{CRAN is the
  largest repository of R code with almost \CranAvailablePackagesRnd packages.
  It receives about 6 new package submissions a day~\cite{Ligges2017}. \cf
  \url{https://cran.r-project.org/web/packages/index.html}}) and the other from
the \emph{Kaggle} data science competition website\footnote{\cf
  \url{https://kaggle.com}}. The intent here is to contrast code written by
experienced developers (CRAN) with code authored by typical end users of the
language (Kaggle). Another difference is that CRAN is a curated repository. Each
submitted package must abide to a number of well-formedness rules that are
automatically checked asserting certain quality. Most relevant for this work is
that all of the runnable code is tested and only a successfully running package
is admitted in the archive. Kaggle on the other hand is more like GitHub, anyone
can submit any code. All the code in our corpus is runnable, \ie we have scripts
that invoke the CRAN packages and the Kaggle programs have input data.

@@

The results presented in this paper are the result of an automated analysis
pipeline that acquires the code of packages, extract metadata, executes
programs, traces their behavior and summarizes the observations.
Figure~\ref{fig:pipeline} shows the main steps of the pipeline along with
approximate time to execute each step, the data size, and the number of elements
manipulated by the stage. Timings are for runs on a cluster of three servers
with Intel Xeon 6140, 2.30GHz with 72 cores and 256GB of RAM using a shared OCFS
network storage.

\begin{figure}[!h]\hspace{-5mm}
  \includegraphics[width=1.05\linewidth]{pipeline.pdf}
  \caption{Analysis Pipeline UPDATE!!}\label{fig:pipeline}
\end{figure}

\noindent The pipeline has the following steps:
\begin{compactenum}
  \item \emph{Download.} The first step consists of downloading and installing
  CRAN packages and acquiring Kaggle corpus. We try to install all available
  CRAN packages. This is somewhat complicated by the fact that many of packages
  depend on native libraries, but these dependencies are not explicitly stated
  in the package description and thus it is hard to automate. For Kaggle we use
  a web crawler to get the programs and the a Kaggle official command line tool
  to download the datasets.
  \item \emph{Runnable code extraction.} Next step is to extract runnabale code.
  For CRAN package we use \genthat~\cite{issta18}, a tool which converts
  runnable code snippets from package documentation and vignettes as well as R
  tests into standalone executable scripts. For Kaggle programs that are not
  already R scripts, we use \c{rmarkdown} and \c{knitr} R packages to extract
  the runnable code from the Rmarkdown or the Jupyter notebook formats
  respectively. Duplicated Kaggle programs are removed. Each of the extracted
  script is wrapped into a call to our dynamic analyzer so we only record eval
  usage triggered by running th script. Without this, the data would include
  eval calls from the unit testing frameworks as well as from bootstrapping R
  virtual machine itself.
  \item \emph{Tracing.} With the scripts ready, we run our dynamic analysis that
  traces the calls to eval and captures other runtime information. For the CRAN
  packages we run two configurations:
  \begin{inparaenum}[(1)]
    \item all of the \CranRunnableScriptsRnd scripts while tracing calls to eval
    originating in any of the \CranAllPackages installed CRAN package, and
    \item a random selection of 1/4 of the scripts while tracing only eval calls
    originating in the core R packages that come bundled with the language.
  \end{inparaenum}
  The reason for this split is that the core R packages are intertwined with the
  language implementation and the usage is different form the user written
  packages hosted on CRAN. For Kaggle, we only trace the eval calls that
  originate in the programs themselves. To avoid any interference, each program
  is run in isolation, in its own R instance.
  \item \emph{Analysis.} In the final step, the output from the dynamic analysis
  is merged, cleaned and summarized in the post-processing phase by custom R
  scripts. The summarized data is then analyzed in RMarkdown notebooks to gather
  insights. Apart from the figures, the data points included in the paper are
  also generated by RMarkdown notebooks as \LaTeX macros.
\end{compactenum}

The code extraction and tracing steps of the pipeline are run in parallel on
three servers using GNU parallel~\cite{GNUparallel}. To make sure, all three
servers have identical environment, we use docker images that have all the
dependencies installed. The pipeline is implemented in 500 lines of Makefile and
3K lines of auxiliary R and shell scripts. Their main function is to gracefully
handle and report failures across large-scale program runs which greatly aids
debugging of the pipeline.

\subsection{Dynamic Analysis}

For this study we need two things: information about calls to eval and related
functions and low-level data such as object allocation and deallocation and
side-effects that happen during eval. The former we can get by injecting code to
the corresponding functions at runtime. For the latter we use
\rdyntrace~\cite{oopsla19a}. It is a modified R virtual machine based on
GNU R 4.0.2 that exposes a number of hooks allowing us to get low-level
information about the R interpreter. \rdyntrace only works with the R
interpreter and not the byte-code compiler. However, this does not affect the
result since the compiler does not optimize eval calls.

We use source location to uniquely identify eval call sites. R only retains
source references at the top-level block expressions. A nested call such as
\c{try(eval(x))} does not have a source reference for the \c{eval(x)}. To work
around this, we impute source locations for eval call sites at runtime, by
traversing functions' AST and attaching source reference attribute manually.

The dynamic analysis is implemented as a R package in 1.5K lines of R code and
2K lines of C++.

\mypara{Limitations.}
%
% We turn off the bytecode compiler for this study. The bytecode
% compiler can also call \eval.
%
We do not get source locations for \UndefinedEvalsRnd UPDATE!! \eval calls. In
these cases \eval is either passed as an argument to a higher-order functions or
is defined in a function returned by a higher-order function, or it can be a
generated called from package native code. However, this is a meager
\PercentUndefinedEval UPDATE!! of all \eval calls and is unlikely to affect our
analysis. We ignore calls to the native \eval function exposed by R. We also
ignore the \c{rlang::tidy_eval} function which uses native \eval internally
because \c{rlang} is used to implement a DSL for data analysis in R. It
introduces a new first-class promise object called \c{quosure} for which it
implements special evaluation support in \c{tidy_eval}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Usage Metrics}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
